A Pilot Multicenter Cluster Randomized Trial to Compare The Effect of Trauma Life Support Training Programs on Patient and Provider Outcomes

### 2022-11-23 pilot-publication/manuscript.Rmd
Draft abstract

**Introduction** Trauma accounts for nearly 10% of the global burden of disease. Several trauma life support programs aim to improve trauma outcomes. There is no evidence from controlled trials to show the effect of these programs on patient outcomes. We describe the protocol of a pilot study that aims to assess the feasibility of conducting a cluster randomised controlled trial comparing Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC) with standard care. **Methods and analysis** We will pilot a pragmatic three-armed parallel, cluster randomised, controlled trial in India, where neither of these programs are routinely taught. We will recruit tertiary hospitals and include trauma patients and residents managing these patients. Two hospitals will be randomised to ATLS, two to PTC, and two to standard care. The primary outcome will be all cause mortality at 30 days from the time of arrival to the emergency department. Our secondary outcomes will include patient, provider, and process measures. All outcomes except time to event outcomes will be measured both as final values as well as change from baseline. We will compare outcomes in three combinations of trial arms: ATLS versus PTC, ATLS versus standard care, and PTC versus standard care using absolute and relative differences along with associated confidence intervals. We will conduct subgroup analyses across the clinical subgroups men, women, blunt multisystem trauma, penetrating trauma, shock, severe traumatic brain injury, and elderly. In parallel to the pilot study we will conduct community consultations to inform the planning of the full-scale trial. **Ethics and dissemination** We will apply for ethics approvals to the local Institutional Review Board (IRB) in each hospital. The protocol will be published to Clinical Trials Registry - India and ClinicalTrials.gov. The results will be published and the anonymized data and code for analysis will be released publicly.

>>>>

**Introduction** Several trauma life support programs aim to improve trauma outcomes but there is no evidence from controlled trials to show the effect of these programs on patient outcomes. We conducted a pilot study to assess the feasibility of conducting a cluster randomised controlled trial comparing the effect of Advanced Trauma Life Support (ATLS) and Primary Trauma Care (PTC) with standard care on patient outcomes. **Methods and analysis** We piloted a pragmatic three-armed parallel, cluster randomised, controlled trial in India, where neither of these programs are routinely taught. We recruited tertiary hospitals and included trauma patients and residents managing these patients. Two hospitals were randomised to ATLS, two to PTC, and three to standard care. The primary outcome was all cause mortality at 30 days from the time of arrival to the emergency department. We conducted community consultations in parallel with the pilot trial. **Ethics and dissemination** We obtained ethical approval from all participating centres. **Results** We included X patients and X residents. The patient recruitment rate was X and the resident recruitment rate was X. The loss to follow up rate was X. The all-cause 30 day mortality was X in the ATLS arm, X in the PTC arm, and X in the standard care arm. **Conclusions** Conducting a full-scale cluster randomised controlled trial comparing the effects of ATLS, PTC, and standard care on patient outcomes should be feasible after incorporating key lessons from this pilot.
